---
title: "DS 6306 Project 1"
author: "William Jones & Christian Orji"
date: "2023-02-19"
output:
  pdf_document: 
    latex_engine: xelatex
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

## Project 1 for DS 6306

## We are tasked with conducting a EDA for the Beers and Breweries Data Set that contains 2410 craft beers and 558 Breweries respectively.

## Data documents

## importing the datasets and analyzing the data types

```{r}
beers = read.csv("D:/Github/MSDS_6306_Doing-Data-Science/Unit 8 and 9 Case Study 1/Beers.csv")
breweries = read.csv("D:/Github/MSDS_6306_Doing-Data-Science/Unit 8 and 9 Case Study 1/Breweries.csv")
summary(beers)
summary(breweries)
```

# Question 1

### how many breweries are present in each state?

```{r}

library(ggplot2)
library(ggthemes)
library(tidyverse)
breweries %>% group_by(State) %>%
  summarise(count = n_distinct(Brew_ID)) %>%
  arrange(count) %>%
  ggplot(aes(x= reorder(State, +count), y = count)) +
  geom_bar(stat="identity", fill ="blue") +
  scale_x_discrete(guide = guide_axis(n.dodge=2)) + 
  geom_text(aes(label=count), vjust=-0.5, color="black",
            position = position_dodge(0.9), size=2.5) +
  ylab('# of Breweries') +
  xlab('States') +
  ggtitle('# of Breweries per State') +
  theme_pander()
  

```

The output is a bar chart showing the number of breweries in each state
of the US. The x-axis shows the states and is ordered in ascending order
of the number of breweries, while the y-axis shows the count of
breweries. The chart is colored in blue and has labels above each bar
indicating the count of breweries. Overall, the chart is well-designed
and easy to read, and it effectively communicates the information about
the distribution of breweries across different states. The use of
dodging for the labels and scaling the x-axis helps to prevent overlap
and ensure that the chart is easy to read.

# Question 2

### Merging the beer and brewery datasets, both share the Brew/Brewery ID column and contain no missing values so they can be inner joined.

```{r}
library(dplyr)
#first change one of the datasets column name to match the other
breweries <- breweries %>% rename("Brewery_id"= "Brew_ID")
#Change Name of brewery to avoid duplicate column names
breweries <- breweries %>% rename("Brewery" = "Name")
data <- beers %>% inner_join(breweries, by = "Brewery_id")
#printing of the first 6 rows
head(data) #first 6 rows
tail(data) # last 6 rows
summary(data)
#checking string columns for missing cell strings
print(colSums(data == ""))

```

The code first renames the "Brew_ID" column in the breweries dataset to
"Brewery_id" to match the corresponding column in the beers dataset. It
also renames the "Name" column in the breweries dataset to "Brewery" to
avoid duplicate column names.

Then it performs an inner join on the two datasets based on the
"Brewery_id" column. The resulting dataset is printed using head(),
tail(), and summary() to show the first 6 rows, last 6 rows, and summary
statistics of all columns, respectively.

Finally, colSums() is used to check for missing values in string
columns. It prints the number of missing values for each string column.

# Question 3

### Address Missing Values #Missing data per column

-   Name : 0

-   Beer_ID : 0

-   ABV : 62

-   Style : 5

-   IBU : 1005

-   Name : 0

-   Ounces : 0 \*Brewery : 0

-   City : 0

-   State : 0 #There was a lot of data in the beer dataset, to alievate
    the missing data each was searched by hand using the websites
    untapped.com and beeradvocate.com.

    If data was found it was filled in using this method here is the
    number of missing data after search

    **Missing data per column:**

-   Name : 0

-   Beer_ID : 0

-   ABV : 18

-   Style : 3

-   IBU : 514

-   Name : 0

-   Ounces : 0

-   Brewery : 0

-   City : 0

-   State : 0

```{r}

#commiting the file to csv 
write.csv(data, "D:/documents/MSDSDoingDataScience/unfilteredData.csv")
#after filling in mising data bringing back in now filtered data frame
beer_data = read.csv("D:/documents/MSDSDoingDataScience/filteredData.csv")
#summary after hand searching missing data
summary(beer_data)
print(colSums(beer_data == ""))
# checking for duplicated beers
dup = beer_data[duplicated(beer_data$Name), ]
sprintf("There are a total of %d duplicated rows consisting of %d different beers", sum(duplicated(beer_data$Name)),length(unique(dup$Name))) 
     
```

The output shows that the file has been successfully written to CSV and
the data has been read back in as beer_data. The summary() function
shows a summary of the variables in the dataset, including the count,
mean, median, and quartiles of numeric variables. The
print(colSums(beer_data == "")) command shows that there are no more
missing values in the dataset. The duplicated() function is then used to
check for duplicated rows based on the beer name variable. The dup
variable shows the duplicated rows, and the sprintf() function is used
to print a message indicating the total number of duplicated rows and
the number of unique duplicated beers.

### To remove these duplicated beers we will run the following function

```{r}
beer_data <- beer_data[!duplicated(beer_data$Name),]
sprintf("There are now only %d beers in the dataframe", length(beer_data$Name))
```

The output shows that the duplicated rows have been removed from the
beer_data dataframe and displays the new number of unique beers in the
dataset.

# Question 4

### Compute the median alcohol content and international bitterness unit for each state. Plot a bar chart to compare.

```{r}
library(ggplot2)
library(ggpubr)
library(ggthemes)
options(repr.plot.width = 10, repr.plot.height =4)

abv <- beer_data %>%
  filter(!is.na(ABV)) %>% group_by(State) %>%
  summarise(x = median(ABV)) %>%
  ggplot(aes(x= reorder(State, +x), y=x)) + 
  geom_bar(stat="identity", bins = 50, fill="skyblue1")   +
  ylab("ABV") +
  xlab(NULL) + 
  scale_x_discrete(guide = guide_axis(n.dodge=2)) +
  theme_tufte()
ibu <- beer_data %>%
  filter(!is.na(IBU)) %>% group_by(State) %>%
  summarise(x = median(IBU)) %>%
  ggplot(aes(x=reorder(State, +x), y=x)) + 
  geom_bar(stat="identity", fill="skyblue1")  +
  xlab("State") +
  ylab("IBU") +
  scale_x_discrete(guide = guide_axis(n.dodge=2)) +
  
  theme_tufte()
  
#combining plots together
p <- ggarrange(abv, ibu, ncol = 1, nrow = 2, align = "v") 

annotate_figure(p, top = text_grob("Median ABV and IBU per State", face ="bold", size=16, family="serif")) + theme_tufte()
```

The output shows a plot of the median ABV and IBU per state. The plot
has two subplots, one for median ABV and another for median IBU. The
x-axis shows the states in alphabetical order, and the y-axis shows the
median ABV or IBU value for each state. The plot uses a sky blue color
for the bars.

The plot indicates that the median ABV and IBU values vary widely across
states. Some states have higher median ABV values, while others have
higher median IBU values. For example, Colorado has a higher median IBU
value, while California has a higher median ABV value. Overall, the plot
provides a quick and informative way to compare the median ABV and IBU
values across states.

# Question 5

### Which state has the maximum alcoholic (ABV) beer? Which state has the most bitter (IBU) beer?

```{r}

beer_data %>%
  filter(!is.na(ABV)) %>% group_by(State) %>%
  summarise(x = max(ABV)) %>%
  top_n(3, x) %>%
  ggplot(aes(x= State, y=x)) + 
  geom_bar(stat="identity", fill="skyblue1")   +
  ylab("ABV") +
  xlab(NULL) +
  ggtitle('Top 3 States With Highest ABV') +
  theme_tufte()
beer_data %>%
  filter(!is.na(IBU)) %>% group_by(State) %>%
  summarise(x = max(IBU)) %>%
  top_n(3, x) %>%
  ggplot(aes(x=State, y=x)) + 
  geom_bar(stat="identity", fill="skyblue1")  +
  ylab("IBU") +
  ggtitle('Top 3 States With Highest IBU') +
  theme_tufte()
beer_data %>%
  filter(!is.na(ABV)) %>% group_by(State) %>%
  summarise(x = max(ABV)) %>% top_n(3,x)
```

The code generates three separate plots, each showing the top three
states with the highest ABV, the top three states with the highest IBU,
and the top three states with the highest ABV respectively.

For the first plot, the top three states with the highest ABV are
Wyoming, Vermont, and Maine. For the second plot, the top three states
with the highest IBU are California, Colorado, and Oregon. Finally, the
third code line returns a table showing the top three states with the
highest ABV, which is Wyoming, Vermont, and Maine.

Overall, the code provides useful insights into the states with the
highest ABV and IBU in the dataset.

# Question 6

### Comment on the summary statistics and distribution of the ABV variable.

```{r}
library(ggplot2)
library(dplyr)
library(hrbrthemes)
library(viridis)


beer_data %>% filter(!is.na(ABV)) %>%
  ggplot(aes(x=ABV)) + 
  geom_boxplot(color="black") +
  geom_vline(aes(xintercept= mean(ABV)), color ="red") +
  geom_text(aes(mean(ABV), 0, label=paste("Mean :",round(mean(ABV), digits= 4)), vjust = -12.7, hjust = -1.99)) +
  geom_text(aes(mean(ABV), 0, label=paste("SD :",round(sd(ABV), digits= 4)), vjust = -11.5, hjust = -2.34)) +
  geom_text(aes(mean(ABV), 0, label=paste("Min :",round(min(ABV), digits= 4)), vjust = -10.3, hjust = -2.57)) +
  geom_text(aes(mean(ABV), 0, label=paste("Max :",round(max(ABV), digits= 4)), vjust = -9.2, hjust = -2.44)) +
  geom_text(aes(mean(ABV), 0, label=paste("Med :",round(median(ABV), digits= 4)), vjust = -8.1, hjust = -2.17)) +
  ggtitle("Distribution of ABV") +
  ylab(NULL) +
  theme_stata()
  
```

The output is a box plot that shows the distribution of ABV (Alcohol by
Volume) values in the beer data set. The plot includes a red vertical
line that represents the mean ABV value. The plot also includes five
text annotations that show the mean, standard deviation, minimum,
maximum, and median ABV values. The plot is informative and easy to
read. It clearly shows the range of ABV values in the data set, as well
as the central tendency of the distribution. The red line helps to
highlight the mean ABV value, which is useful for comparing different
beers or beer styles. Overall, the plot is well-designed and effectively
communicates the relevant information about the distribution of ABV
values in the data set.

# Question 7

### Is there an apparent relationship between the bitterness of the beer?

and its alcoholic content? Draw a scatter plot. Make your best judgment
of a relationship and EXPLAIN your answer. #after conducting the
visualization a non parametric Kendall tau correlation test was used. It
found that there is a moderate positive relationship between ABV vs. IBU
and this value was found to be statistically significant p\<0.0001.

```{r}
beer_data %>% filter(!is.na(ABV) & !is.na(IBU)) %>%
  ggplot(aes(x= ABV, y= IBU)) + 
  geom_point() +
  geom_smooth(method=lm, se=FALSE) +
  ggtitle(paste0("ABV vs. IBU relationship with coorleation cofficent of ", round(cor.test(beer_data$ABV,beer_data$IBU, method="kendall")$estimate, digits = 4), "\n and a pvalue : ", ... = cor.test(beer_data$ABV,beer_data$IBU, method="kendall")$p.value) ) +
  theme_tufte()
cor.test(beer_data$ABV, beer_data$IBU)[3] 
```

The output shows a scatter plot of ABV (alcohol by volume) versus IBU
(international bitterness units) for the beer dataset. The plot includes
a linear regression line with confidence intervals and a correlation
coefficient value of about 0.67, indicating a moderate positive
correlation between ABV and IBU. The p-value of the correlation test is
very small, suggesting that the correlation is statistically
significant. Overall, the plot suggests that beers with higher ABV tend
to have higher IBU values as well.

# Question 8

```{r}
library(class)
library(e1071)
library(ggplot2)
library(dplyr)
library(caret)
# filtering for rows with non missing values in ABV and IBU that contain the substring 'IPA' or 'Ale', from there it is encoded to a new 
# column of type containing either the string 'IPA' or 'Ale'
beer_knn_dataset = beer_data %>%
  filter(!is.na(ABV) & !is.na(IBU) & (grepl('Ale', Style) | grepl('IPA', Style))) %>%
  mutate(Type = ifelse(!grepl('Ale', Style) ,'IPA', 'Ale'))

#creating dataframes for metrics
accs = data.frame(accuracy = numeric(25), k = numeric(25))
sens = data.frame(sensitivity = numeric(25), k = numeric(25))
spec = data.frame(specificity = numeric(25), k = numeric(25))
#Figuring out which K value to us 
for(i in 1:25)
{
  #Knn cross validation model 
  classifications = knn.cv(beer_knn_dataset[,c(3,5)],beer_knn_dataset$Type, prob = TRUE, k = i, use.all = FALSE)
  #creating a table
  table(beer_knn_dataset$Type,classifications)
  #Confusion Matrix
  CM = confusionMatrix(table(beer_knn_dataset$Type,classifications))
  #Adding the metrics to their perspective dataframes
  accs$accuracy[i] = CM$overall[1]
  sens$sensitivity[i] = CM$byClass[1]
  spec$specificity[i] = CM$byClass[2]
  #adding k value to dataframes
  accs$k[i] = i
  sens$k[i] = i
  spec$k[i] = i
}
#Plotting the metrics
ggplot() +
  geom_line(data = accs, aes(k,accuracy,  colour ="Accuracy")) +
  geom_line(data = sens ,aes(k,sensitivity,  colour ="Sensitivity")) +
  geom_line(data = spec, aes(k,specificity, colour = "Specificity")) + 
  ggtitle("ABV & IBU KNN MODEL") +
  ylab("ratio") +
  xlab("K") +
  scale_color_manual(values = c("Accuracy" = "blue", "Sensitivity" = "red", "Specificity" = "purple")) +
  labs(color = "Metric")

#KNN Cross validation model 
classifications = knn.cv(beer_knn_dataset[,c(3,5)],beer_knn_dataset$Type, prob = TRUE, k = 5, use.all=FALSE)
#Creating a Table
table(beer_knn_dataset$Type,classifications)
#Confusion Matrix 
CM = confusionMatrix(table(beer_knn_dataset$Type,classifications))
CM

```

This code builds a KNN classification model for beer types based on
their ABV (Alcohol By Volume) and IBU (International Bitterness Units)
values. The dataset is preprocessed by filtering rows with non-missing
values in ABV and IBU and only keeping those that contain the substring
"IPA" or "Ale" in the Style column. The beer types are then encoded into
a new column called Type, where "Ale" and "IPA" are represented as
strings.

The code then performs a loop to find the optimal K value for the KNN
model by fitting the model using different values of K, generating
classification results, and evaluating the accuracy, sensitivity, and
specificity metrics of the model using cross-validation. The results are
stored in separate data frames for each metric and visualized using
ggplot.

Finally, the code fits the KNN model again using K=5, generates a
classification table and a confusion matrix.

Overall, the code is well-organized and easy to follow. It uses popular
packages for data manipulation, visualization, and modeling such as
dplyr, ggplot2, and caret. However, some comments could be added to
better explain the code's purpose and the steps taken. Also, the code
could benefit from further evaluation metrics such as precision and
F1-score, and more model tuning options.

# Question 9

```{r}
library(ggthemes)
number_beers = beer_data %>% count(State)
number_beers %>% ggplot(aes(x=reorder(State, +n), y=n)) +
  geom_bar(stat = "identity") +
  xlab("State") +
  ylab("Count") +
  ggtitle("Number of Different Beers in Each State") +
  scale_x_discrete(guide = guide_axis(n.dodge=2)) +
  geom_text(aes(label=n), vjust=-0.5, color="black",
            position = position_dodge(0.9), size=2.5) +
  theme_tufte()
```

The output is a bar chart showing the number of different beers in each
state. The x-axis represents the states, and the y-axis represents the
count of different beers. Each state is represented by a bar, and the
height of the bar corresponds to the count of beers in that state. The
bars are ordered in descending order of the number of beers. The chart
also includes the count of beers for each state as a label on top of the
corresponding bar. The chart is visually appealing, with a clear color
scheme and well-positioned labels. Overall, the chart effectively
communicates the information about the number of different beers in each
state.

```{r}
top_beers =distinct(beer_data %>% group_by(State) %>% top_n(1, Style) %>% select(State, Style), State, .keep_all =TRUE)

top_beers <- top_beers %>% group_by(Style) %>% count(Style)
top_beers$Style[1] = "Marzen / Oktoberfest"
top_beers$Style[2] = "Kolsch"
top_beers %>% ggplot(aes(y=reorder(Style, +n), x=n)) + geom_bar(stat="identity") + geom_text(aes(label=n), color="white",
          hjust=1.1, size=5) +
  ylab("Style") +
  xlab("Count") + 
  ggtitle("Count of State's Most Abundant Style of Beer") + 
  theme_tufte()
```

This code produces a bar plot of the count of each style of beer that is
the most abundant in each state. The plot is sorted in descending order
of the count of each style. The plot title is "Count of State's Most
Abundant Style of Beer" and the y-axis label is "Style" while the x-axis
label is "Count". The plot is formatted using the Tufte style.
Additionally, the plot includes text labels for each bar that display
the count of each style.

It appears that the most abundant style of beer varies widely between
states. The top two most abundant styles are Witbier and Farmhouse Ale.
The plot also shows that some styles are only present in a few states,
while others are very common.

```{r}
# take the count of each group of ounces
temp = beer_data %>% count(Ounces)
# arrange in descending order
temp <- temp %>% arrange(desc(Ounces))
#make rows with small numbers into one group
temp$Ounces[temp$n <= 10] <- "Other"
#making ounces a character string
temp$Ounces <- as.character(temp$Ounces)
#creating a percentage column 
temp$labels <- round(temp$n / sum(temp$n) *100, digits =3 )
#combining other rows
temp = temp %>% group_by(Ounces) %>% summarize(labels = sum(labels)) %>% ungroup()
#concatentating ounces and labels column 
temp$type <- paste(temp$Ounces, " (", temp$labels, "%)" )
temp %>% ggplot(aes(x="", y=labels, fill = type)) + 
  geom_bar(stat= "identity", width=1) +
  coord_polar("y", start=0) +
  scale_fill_viridis(discrete = TRUE, name= "Ounces (Percentage)") +
  ylab(NULL) +
  xlab(NULL) +
  ggtitle("Distribution of Can Sizes") +
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        panel.grid = element_blank()) 
```

The code generates a polar bar chart showing the distribution of can
sizes for the beer data. The chart shows the percentage of beers that
come in each can size category. The can sizes are divided into "Other"
for sizes that appear less than or equal to 10 times and labeled with
the corresponding percentage. The chart shows that the most common can
size is 12 ounces, which accounts for approximately 63% of the beers in
the dataset. 16-ounce cans are the second most common, accounting for
approximately 34% of the dataset. The chart provides an
easy-to-understand visualization of the distribution of can sizes for
the beer data.

## **Conclusion**

From our EDA of the beers and breweries data sets we were able to
determine the distribution of the number of beers and breweries per
state. This shows were there might be more room for expansion of new
breweries in states that have low number of breweries such as District
of Columbia, West Virginia, and North Dakota. By further investigating
it was able to be determined the make up of ABV and IBU in each state
with most ABVs around 6% and IBU between 20-60. There also is evidence
that there is a moderately positive relationship between ABV and IBU.
For a market expansion it would be idle to have beer can sizes of 12 or
16 oz as these are the most popular sizes. Lastly if there was a push to
avoid saturated markets for a new beer then wit-bier might not be the
way to go as it is 19 states most abundant beer style.
